A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-06 tex/BiSig.tex
[ tex ] Rename `mylhs2tex` to `agda-lhs2tex`
[ tex ] Fix the dangling end in abstract

\usepackage{microtype}
\usepackage{subfiles}

@@

\input{type-notation} %% for type theory notation copied from HoTT Book
\input{macros}
\usepackage{mylhs2tex}

@@

\begin{abstract}
%  Type checking and inference serve as the transition from parsed abstract syntax trees to well-typed programs.
  While there is much progress in designing bidirectional type systems and associated type synthesis algorithms on a case-by-case basis, the lack of a general theory makes it difficult to study the common principles and develop widely applicable techniques and practical tools such as a `type-synthesiser generator'.
  In response to this difficulty, we propose a simple yet general and constructive theory of bidirectional type synthesis.
%  which refers to the mutual use of type checking and synthesis based on bidirectional typing rules to determine the type of a program with type annotations.
  First, our theory is simple yet general as it applies to any syntax-directed bidirectional simple type system that can be specified by a signature.
%  which is analogous to a grammar in parsing.
  Second, our theory is constructive because it is not only formalised in a proof assistant based on Martin-L\"of type theory, but also formulated in a way that unifies computation and proof, resulting in a compact and practical theory---for example, a proof of the decidability of the type synthesis problem is also computationally a type synthesiser.
  Moreover, we introduce `mode preprocessing' to deal with missing type annotations, an aspect often overlooked in the literature.
  The main theorem is logically a trichotomy on parsed syntax trees and computationally a type synthesiser that may throw an exception due to missing type annotations.
  Being parametrised by a signature, our theory works as a verified type-synthesiser generator overall.
\end{abstract}

@@

\keywords{bidirectional typing, type synthesis, type checking, type inference, language formalisation, \Agda}

>>>>

\usepackage{microtype}
\microtypesetup{babel}
\usepackage{subfiles}

@@

\input{type-notation} %% for type theory notation copied from HoTT Book
\input{macros}
\usepackage{agda-lhs2tex}

@@

\begin{abstract}
%  Type checking and inference serve as the transition from parsed abstract syntax trees to well-typed programs.
  While there is much progress in designing bidirectional type systems and associated type synthesis algorithms on a case-by-case basis, the lack of a general theory makes it difficult to study the common principles and develop widely applicable techniques and practical tools such as a `type-synthesiser generator'.
  In response to this difficulty, we propose a simple yet general and constructive theory of bidirectional type synthesis.
%  which refers to the mutual use of type checking and synthesis based on bidirectional typing rules to determine the type of a program with type annotations.
  First, it is simple yet general as it applies to any syntax-directed bidirectional simple type system that can be specified by a signature.
%  which is analogous to a grammar in parsing.
  Second, it is constructive because it is not only formalised in a proof assistant based on Martin-L\"of type theory, but also formulated in a way that unifies computation and proof, resulting in a compact and practical theory---for example, a proof of the decidability of the type synthesis problem is computationally a type synthesiser.
  Moreover, we introduce `mode preprocessing' to deal with missing type annotations, an aspect often overlooked in the literature.
  The main theorem is logically a trichotomy on parsed syntax trees and computationally a synthesiser that may throw an exception due to missing type annotations.
  By being parametrised by a signature, our theory works as a verified type-synthesiser generator overall.
\end{abstract}

@@

\keywords{type checking, type inference, language formalisation, \Agda}
